\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[includehead, headheight=10mm, margin=15mm ]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{titling}
\usepackage{fancyhdr}
\usepackage{listings}


\title{APPM 4600 Lab 4}
\author{Edward Wawrzynek}
\date{19 September 2024}

\newcommand*{\dif}{\mathop{}\!\mathrm{d}}

\makeatletter
\def\@maketitle{%
  \newpage
  \null
  \vskip 1em%
  \begin{center}%
  \let \footnote \thanks
    {\LARGE \@title \par}%
    \vskip 1em%
    {\normalfont \@date}
  \end{center}%
  \par
  \vskip 1em}
\makeatother

\begin{document}

\pagestyle{fancy}
    \fancyhf{} % clear all header and footer fields
    \fancyhead[L]{\thetitle}
    \fancyhead[R]{\theauthor}

\makeatletter
\begin{center}
    {\Large \@title}
    \vskip 1mm
    {\normalfont \@date}
    \vskip 1em
\end{center}
\makeatother

\section{Prelab}
\begin{enumerate}
\item Code for order determination is on github and listed at the end of this document in the function \texttt{order\_compare}.
\item \begin{enumerate}
  \item It takes 12 iterations for fixed point to converge onto the fixed point.
  
  \item We find that \begin{align*}
      \alpha &\approx 0.99621, \\
      \lambda  &\approx 0.12365.
  \end{align*} This makes sense, as we expect the method to have first order convergence.
\end{enumerate}
\end{enumerate}

\section{Aitken's $\Delta^2$ method}
\begin{enumerate}
  \item We have that \begin{align*}
      \frac{p_{n+1} - p}{p_n - p} = \frac{p_{n+2} - p}{p_{n+1} - p},
  \end{align*} which we can solve as \begin{align*}
      0 &= (p_{n+1}-p)^2 - (p_{n+2} - p)(p_n-p) \\
      &= p_{n+1}^2 - 2pp_{n+1} + p^2 - p_{n+2}p_n + pp_{n+2} + pp_n - p^2 \\
      &= p_{n+1}^2 - p_{n+2}p_n + p(-2p_{n+1} + p_{n+2} + p_n),
  \end{align*} thus, \begin{align*}
    p &= \frac{p_{n+1}^2 - p_{n+2}p_n}{p_{n+2} - 2p_{n+1}+p_n} \\
    &= p_n - \frac{(p_{n+1}-p_n)^2}{p_{n+2} - 2p_{n+1}+p_n},
  \end{align*} as expected.

  \item When we apply Aitken's method to the sequence that we found previously, we converge in 5 iterations and find that \begin{align*}
    \alpha &\approx 1.303 \\
    \lambda &\approx 0.835,
  \end{align*} which is super linear convergence (\(\alpha > 1\)), faster than fixed point iteration.
\end{enumerate}

\section{Steffenson's method}
\begin{enumerate}
\item The implementation of steffenson's method is in the code below and on github.

\item Steffensons's method yields the root \begin{align*}
    r \approx 1.36523001341409,
\end{align*} as expected.

\item Steffenson's method converges with \begin{align*}
    \alpha &\approx 1.996,
    \lambda &\approx 0.001922
\end{align*} after 3 iterations. Notice that this appears to be quadratic convergence, which is faster than un modified fixed point's linear convergence.
\end{enumerate}

{\small \lstinputlisting[language=Python]{lab4.py}}   

\end{document}